\textbf{COMPARISON OF \\ HEURISTIC ALGORITHMS \\ FOR KP01} \\[0.5cm] \rule{12cm}{0.3mm} \\[0.5cm] \small \scshape{A comparison of heuristic algorithms for solving the 0-1 Knapsack Problem} \\[0.4cm] \rule{12cm}{0.3mm} \vskip -0.3cm

### 2023-04-13 doc/paper.tex
rough finish of harmony alg in doc

\begin{algorithm}
\caption{Algorithm 2: Solving 0-1 Knapsack with Dynamic Programming}\label{dp}
    \begin{algorithmic}
        \For {$i = 0 \text{ to } noItems $} \Comment{// If no items, then profit = 0}
            \State $Table[i][0] \gets 0 $ 
        \EndFor
        \For {$k = 0 \text{ to } maxCapacity $} \Comment{// If no capacity, then profit = 0}
            \State $Table[0][k] \gets 0 $
        \EndFor
        \For {$i = 0 \text{ to } noItems$}
            \For {$k = 0 \text{ to } maxCapacity $}
                \State $Table[i][k] \gets Table[i-1][k]$  
                \If {$k-weights[i] >= 0$} 
                    \State $Table[i][k] \gets max(Table[i-1][k-weights[i]] + profits[i], Table[i][k]) $
                \EndIf
            \EndFor
        \EndFor
        \State $\textbf{Print } Table[noItems][maxCapacity] $
    \end{algorithmic}
\end{algorithm}

@@

It consists of 4 main parts:
\begin{enumerate}
    \item Create a harmony memory HM - a fixed size (HMS) of randomly generated harmonies. Initialize parameters harmony memory consider rate (HMCR), and pitch adjusting rate (PAR), and calculate profit-density vector for usage in the repair-operator.
    \item Create a new harmony using the current HM (\textcolor{red}{Algorithm x}).
    \item If generated harmony is better than the best in the harmony, replace it. If not, check if it is better than then worst harmony, and replace it if so. 
    \item If maximum number of iterations has been met, output the best sum profit value found so far.
\end{enumerate}

@@

\begin{algorithm}
\caption{Generating a new harmony during iterative part (part 2)}\label{harmonyGen}
    \begin{algorithmic}
        \For {$i = 1 \text{ to } noItems$}
            \If {$rand(0, 1) <= HMCR(t)$}
                \State $newHarmony[i] \gets bestHarmony[i]$ 
            \Else
                \State Generate a random integer number $a \in \{1, 2, ... HMS\}, a \neq best$
                \State $newHarmony[i] \gets Harmony_{a}[i]$ 
                \If {$rand(0, 1) <= PAR(t)$}
                    \State $newHarmony[i] = |newHarmony[i]-1|$ \Comment{// Flipping i'th item - mutation}
                \EndIf
            \EndIf
        \EndFor
    \end{algorithmic}
\end{algorithm}

Any time a harmony is generated, during Part 1 or Part 2, it is "repaired" in case the selection of items exceeds the capacity of the knapsack. The new generated harmony is also repaired if it can fit more items, without exceeding the capacity of the knapsack. The repair-operator consists of 2 phases:
\begin{enumerate}
    \item Drop phrase - repairing a harmony if it violates the constraint
    \item Add phase - adding a items into the knapsack if the total weight is less than the capacity
\end{enumerate}

@@

\begin{algorithm}
\caption{Repair-operator after harmony generation}
    \begin{algorithmic}
        \State \Comment{// Profit-density vector precomputed during Part 1 (initialization)}
        \For {$i = 1 \text{ to } N$}
            \State $\lambda_{i} = \frac{profit_{i}}{weight_{i}}$
        \EndFor
        \State \Comment{Actual repair-operator starts here}
        \If {$totalWeight > maxWeight$}
        \State \Comment{// DROP phase}

        \EndIf
    \end{algorithmic}
\end{algorithm}

@@

\paragraph{Pseudorandom numbers}

@@

First, the number of solutions in memory (for the heuristic algorithms) was put in the input file (20 in all testcases and algorithms for this report). In all testcases, the number of items $n$ was chosen to either be 1000 or 100000. The profit and weight values were then randomly generated in a range of \, $ [1, n] $ while a sum-of-weights variable \emph{sum} was kept. The max weight was then randomly generated in a range of $[\frac{\text{sum}}{100} , \frac{\text{sum}}{10}]$, so that the knapsack wouldn't be able to hold all the items. To avoid any overflow errors, all generated numbers (and the sum of the profit/weight values) were kept below the max value of an \emph{int} which is $2^{31}-1$ in many programming languages.  Then $n$, $W$ (maxWeight), and each item (profit and weight values) was put in an input file.

>>>>

\begin{breakablealgorithm}
\caption{Solving 0-1 Knapsack with Dynamic Programming}\label{dp}
    \begin{algorithmic}[1]
        \For {$i = 0 \text{ to } noItems $} \Comment{\textcolor{blue}{If no items, then profit = 0}}
            \State $Table[i][0] \gets 0 $ 
        \EndFor
        \For {$k = 0 \text{ to } maxCapacity $} \Comment{\textcolor{blue}{If no capacity, then profit = 0}}
            \State $Table[0][k] \gets 0 $
        \EndFor
        \For {$i = 0 \text{ to } noItems$}
            \For {$k = 0 \text{ to } maxCapacity $}
                \State $Table[i][k] \gets Table[i-1][k]$  
                \If {$k-weights[i] >= 0$} 
                    \State $Table[i][k] \gets max(Table[i-1][k-weights[i]] + profits[i], Table[i][k]) $
                \EndIf
            \EndFor
        \EndFor
        \State $\textbf{Print } Table[noItems][maxCapacity] $
    \end{algorithmic}
\end{breakablealgorithm}

@@

It consists of 4 main parts:
\begin{enumerate}
    \item Create a harmony memory HM - a fixed size (HMS) of randomly generated harmonies. Initialize parameters harmony memory consider rate (HMCR), and pitch adjusting rate (PAR), and calculate profit-density vector for usage in the repair-operator.
    \item Create a new harmony using the current HM \ref{harmonyGen}.
    \item If generated harmony is better than the best in the harmony, replace it. If not, check if it is better than then worst harmony, and replace it if so. 
    \item If maximum number of iterations has been met, output the best sum profit value found so far.
\end{enumerate}

\begin{breakablealgorithm}
\caption{The DGHS algorithm}\label{DGHS}
    \begin{algorithmic}[1]
        \State Set the harmony memory size $HMS$, the number of iterations $ITERATIONS$, and the minimum and maximum values of parameters $PAR$ and $HMCR$.
        \State Initialize the HM, and use \ref{harmonyRepair} to the generated harmonies. Calculate the totalProfit and totalWeight values for each harmony in HM.
        \State $iterator \gets 1$
        \While {$iterator <= ITERATIONS$}
            \State Record indices of the best and the worst harmonies in HM.
            \State Calculate parameters HMCR and PAR for the current iteration.
            \State Perform Algorithm \ref{harmonyGen} to produce a new harmony $\Vec{x}_{new}$
            \State Perform Algorithm \ref{harmonyRepair} to repair the new harmony $\Vec{x}_{new}$
            \If {$\Vec{x}_{new} \text{ is better than or equal to } \Vec{x}_{best}$}
                \State Replace $\Vec{x}_{best} \text{ with } \Vec{x}_{new}$
            \ElsIf {$\Vec{x}_{new} \text{ is better than or equal to } \Vec{x}_{worst}$}
                \State Replace $\Vec{x}_{worst} \text{ with } \Vec{x}_{new}$
            \EndIf
            \State $iterator \gets iterator+1$
        \EndWhile
        \State Output profit of best harmony
    \end{algorithmic}
\end{breakablealgorithm}

@@

\begin{breakablealgorithm}
\caption{Generating a new harmony during iterative part (part 2)}\label{harmonyGen}
    \begin{algorithmic}[1]
        \For {$i = 1 \text{ to } noItems$}
            \If {$rand(0, 1) <= HMCR(t)$}
                \State $newHarmony[i] \gets bestHarmony[i]$ 
            \Else
                \State Generate a random integer number $a \in \{1, 2, ... HMS\}, a \neq best$
                \State $newHarmony[i] \gets Harmony_{a}[i]$ 
                \If {$rand(0, 1) <= PAR(t)$}
                    \State $newHarmony[i] = |newHarmony[i]-1|$ \Comment{// Flipping i'th item - mutation}
                \EndIf
            \EndIf
        \EndFor
    \end{algorithmic}
\end{breakablealgorithm}

Any time a harmony is generated, during Part 1 or Part 2, it is "repaired" in case the selection of items exceeds the capacity of the knapsack. The new generated harmony is also repaired if it can fit more items, without exceeding the capacity of the knapsack. The repair-operator consists of 2 phases:
\begin{enumerate}
    \item Drop phase - repairing a harmony if it violates the constraint
    \item Add phase - adding a items into the knapsack if the total weight is less than the capacity
\end{enumerate}

@@

\begin{breakablealgorithm}
\caption{Repair-operator after harmony generation}\label{harmonyRepair}
    \begin{algorithmic}[1]
        \If {$totalWeight > maxWeight$}
            \For {$i = 1 \text{ to } N$} \Comment{ DROP phase}
                \State $\lambda_{i} = \frac{profit_{i}}{weight_{i}}$
            \EndFor
            \State Sort items in increasing order of $\lambda_{i}$, and let $ind_{i}$ denote the original index of each $\lambda_{i}$
            \For {$i = 1 \text{ to } noItems $} 
            \State Remove the ones with the least profit-density values greedily
                \If {$\lambda_{i} == 0$}
                    \State \textbf{Continue}
                \EndIf
                \State $newHarmony[ind_{i}] = 0$ \Comment{Unload the item}
                \State $totalWeight \gets totalWeight - weight[ind_{i}]$
                \State $totalProfit \gets totalProfit - profit[ind_{i}]$
                \If {$totalWeight <= maxWeight$}
                    \State \textbf{Break} \Comment{Terminate DROP phase}
                \EndIf
            \EndFor
        \EndIf
        \If {$totalWeight < maxWeight$}
            \For {$i = 1 \text{ to } N$} \Comment{ADD phase}
                \State $\lambda_{i} = \frac{profit_{i}}{weight_{i}}$
            \EndFor
            \State Sort items in increasing order of $\lambda_{i}$, and let $ind_{i}$ denote the original index of each $\lambda_{i}$
            \For {$i = 1 \text{ to } noItems $} 
                \State Add the ones with the greatest profit-density if possible 
                \If {$newHarmony[ind_{i}] == 0$}
                    \If {$totalWeight + weight[ind_{i}] <= maxWeight$}
                        \State $newHarmony[ind_{i}] = 1$
                        \State $totalWeight \gets totalWeight + weight[ind_{i}]$
                        \State $totalProfit \gets totalProfit + profit[ind_{i}]$
                    \EndIf
                \EndIf
            \EndFor
        \EndIf
    \end{algorithmic}
\end{breakablealgorithm}

@@

@@

First, the number of solutions in memory (for the heuristic algorithms) was put in the input file which was kept constant, 20, in all testcases and algorithms for this report. In all testcases, the number of items $n$ was chosen to either be 1000 or 100000. The profit and weight values were then randomly generated in a range of \, $ [1, n] $ while a sum-of-weights variable \emph{sum} was kept. The max weight was then randomly generated in a range of $[\frac{\text{sum}}{100} , \frac{\text{sum}}{10}]$, so that the knapsack wouldn't be able to hold all the items. To avoid any overflow errors, all generated numbers (and the sum of the profit/weight values) were kept below the max value of an \emph{int} which is $2^{31}-1$ in many programming languages.  Then $n$, $W$ (maxWeight), and each item (profit and weight values) was put in an input file.
